\chapter*{}
\begin{center}
\label{abstract}
\textbf{Abstract}
\end{center}
Traditional blockchain education often prioritizes theoretical concepts, leaving a significant gap in the technical expertise required for real-world implementation. This thesis addresses this deficiency by presenting a modular educational blockchain system and an accompanying educational curriculum designed for university-level computer science students. The proposed solution features a Python-based blockchain system that facilitates a hands-on learning path, progressing from basic networking to the implementation of consensus mechanisms. The curriculum further requires students to benchmark system performance, analyzing critical metrics such as throughput and energy efficiency. To illustrate the versatility of blockchain technology beyond cryptocurrencies, the thesis explores non-financial applications, exemplified by using blockchains as a digital asset manager during the educational curriculum. The practical component is complemented by a Vue.js-based blockchain explorer, offering real-time visualization of internal network states, including mempool data and block propagation. This work aims to provide students with an educational ecosystem, combining a functional software solution with a structured educational framework to enhance student understanding of distributed systems and blockchains.\vspace{0.3cm}

\textbf{Keywords:} Blockchain Education, Cryptocurrency Prototype, Laboratory Curriculum, Consensus Algorithms, Proof of Stake, Blockchain Explorer, Python.