\chapter{Conclusion and Future Work} \label{ch7_conclusion}
\markboth{Conclusion and Future Work}{Conclusion and Future Work}

This thesis set out to address the widening gap between the theoretical principles of distributed systems and the practical engineering skills required to build them. By designing a modular, Python-based blockchain system and a corresponding constructionist laboratory curriculum, this work provides a comprehensive educational framework that transforms students from passive consumers of technology into active system architects. This final chapter summarizes the primary contributions of the research and outlines strategic directions for future development.

\section{Summary of Contributions} \label{sec:contributions}
The contributions of this thesis are twofold, advancing both the software tooling available for blockchain education and the educational methodologies used to teach it.

\subsection{The Educational Blockchain Artifact}
The primary technical contribution is the development of a "Glass-Box" blockchain node designed specifically for educational transparency. Unlike production systems such as Bitcoin or Ethereum, which prioritize performance and security at the cost of complexity, this system prioritizes readability and modularity. Key technical achievements include:
\begin{itemize}
    \item \textbf{Modular Consensus Engine:} The implementation of the Strategy Pattern allows for the dynamic runtime swapping of consensus algorithms. This enabled the direct empirical comparison of Proof of Work (PoW) and Proof of Stake (PoS) within a controlled environment, isolating the consensus variable from network noise.
    \item \textbf{Containerized Network Simulation:} The integration of Docker orchestration allows students to deploy complete P2P networks on personal hardware. This lowers the barrier to entry for studying distributed phenomena like propagation latency and network partitions.
    \item \textbf{Visual Observability:} The development of a Vue.js-based Blockchain Explorer provides students with immediate visual feedback on the state of the ledger, reinforcing their mental model of the data structures.
\end{itemize}

\subsection{The Laboratory Curriculum}
The educational contribution is a structured 5-Module Curriculum that guides students through the engineering lifecycle of a distributed system. Validated through a pilot study, the curriculum effectively targets the "Zone of Proximal Development" for computer science students:
\begin{itemize}
    \item \textbf{Scaffolding Complex Concepts:} By providing "Starter Code" that abstracts low-level socket management (Module 1), students can focus on high-level logic.
    \item \textbf{Empirical Analysis:} Module 3 successfully shifts the focus from "getting it to work" to "measuring how well it works," instilling a rigourous engineering mindset regarding benchmarking and metrics (TPS, Latency).
    \item \textbf{Architectural Criticism:} Module 5 forces students to confront the limitations of their own creations, deriving architectural best practices (e.g., hybrid off-chain storage) through the failure analysis of naive implementations.
\end{itemize}

\section{Future Work} \label{sec:future_work}
While the current system successfully meets its educational objectives, the pilot study identified specific areas where the framework can be evolved to improve both technical fidelity and the student learning experience.

\subsection{educational Refinement}
The evaluation in Chapter \ref{ch6_evaluation} revealed that the learning curve for implementing consensus mechanisms (Module 2) and system extensibility (Module 5) was steeper than anticipated. Future iterations of the curriculum should introduce intermediate "bridging" tasks. Specifically, the difficulty adjustment algorithm in Proof of Work proved to be a significant stumbling block. Future course materials should provide more granular theoretical derivations and pseudo-code for the retargeting logic before requiring students to implement it in Python. Additionally, enhanced debugging tools and pre-written unit tests for the consensus module would prevent students from getting blocked by minor syntax errors, allowing them to focus on the algorithmic logic.

\subsection{Runtime Environment and Language Migration}
A critical technical limitation identified in this work is the performance bottleneck imposed by the Python Global Interpreter Lock (GIL). While Python's expressiveness makes it ideal for educational readability, the GIL prevents true parallel execution, limiting the fidelity of race-condition simulations and capping throughput at approximately 16 TPS.

To address this, future work should explore migrating the core node implementation to a language that supports true multi-threading while maintaining a manageable cognitive load for students. 
\begin{itemize}
    \item \textbf{Golang (Go):} Go offers native concurrency primitives (goroutines) that map well to distributed systems concepts. Its syntax is simpler than C++, making it a viable candidate for upper-level undergraduate courses.
    \item \textbf{Kotlin:} As a modern JVM language, Kotlin offers high performance and robust concurrency (coroutines) while remaining syntactically approachable for students familiar with Java or Python.
\end{itemize}
Transitioning to either language would represent a compromise: gaining system performance and parallelism fidelity at the cost of a higher barrier to entry for students with limited systems programming experience.

\subsection{Advanced Networking and Smart Contracts}
The current P2P layer utilizes a simple broadcast gossip protocol. To better reflect the complexity of modern production networks, future versions of the system could implement a \textbf{Kademlia Distributed Hash Table (DHT)} for peer discovery. This would allow the curriculum to cover advanced topics such as routing efficiency and sybil resistance in the networking layer.

Furthermore, the "Smart Assets" module (Module 4) currently relies on hard-coded transaction logic. Extending the system to include a lightweight virtual machine would allow for the introduction of Turing-complete \textbf{Smart Contracts} (e.g., a simplified Solidity subset). This would significantly expand the scope of the curriculum to include decentralized finance (DeFi) primitives and automated governance, aligning the educational content with the current state of the industry.