%%%%% Questions?
%%%%%%%%%%%%%%%%%%%%%% 
%%%%%%%%%%%%%%%%%%%%%% 

\chapter{Results and Discussion} \label{ch6_evaluation}
\markboth{Results and Discussion}{Results and Discussion}

Following the design and implementation of the modular blockchain system in Chapter \ref{ch4_implementation} and the educational curriculum in Chapter \ref{ch5_curriculum}, this chapter presents the evaluation of the proposed solution. The core objective of this thesis was to bridge the gap between theoretical concepts and practical engineering in distributed systems education. Therefore, the evaluation strategy must assess both the technical robustness of the software artifact and the feasibility of the educational material.

The evaluation is conducted in two distinct phases:

\begin{enumerate}
    \item \textbf{Phase I: Technical System Evaluation.} This phase involves quantitative benchmarking of the system. By subjecting the blockchain node to high-load stress tests (as defined in Module 3 \ref{ch5_module3}), we analyze critical metrics such as Transaction Throughput (TPS), Block Propagation Latency, and Resource Utilization.  The objective of this phase is to verify that the educational blockchain system faithfully reproduces the practical constraints and performance trade-offs observed in real-world blockchain platforms, particularly those associated with the Scalability Trilemma, as exemplified by production systems such as Bitcoin and Ethereum \cite{buterintrilemma}.
    
    \item \textbf{Phase II: Curriculum Validation.} This phase involves a qualitative evaluation and assessment of the instructional effectiveness of the educational curriculum in Chapter \ref{ch5_curriculum}. A preliminary pilot study was conducted with a cohort of participants with Computer Science backgrounds. This phase focuses on evaluating the clarity of the instructional materials, the success rate of task completion, and the overall usability and maintainability of the provided codebase.
\end{enumerate}

This dual-approach ensures that the system is not only technically sound but also accessible for its intended educational purpose.